---
title: "Time Series Analysis of Monthly Drinking Milk Production in Portugal"
output: html_notebook
---

## 1. Introduction

*(to be written)*

---

## 2. Data and Exploratory Analysis

### 2.1 Packages

Necessary packages

```{r message=FALSE, warning=FALSE}
#install.packages(c("readxl","lubridate","dplyr","ggplot2","forecast","tseries","urca","lmtest","MuMIn","kableExtra"))

library(readxl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(urca)
library(lmtest)
library(MuMIn)
library(kableExtra)
```

### 2.2 Data Loading and Preparation

```{r}
dados <- read.csv("data/leite.csv")
dados
```

Select the relevant columns and remove the three empty rows at the start (1986, no data reported).




```{r}
dadosSelected <- select(dados, TIME_PERIOD, OBS_VALUE)

dadosSelected <- dadosSelected[-c(1,2, 3), ]

# Ensure numeric values (handles any locale comma decimals)
dadosSelected$OBS_VALUE <- as.numeric(gsub(",", ".", dadosSelected$OBS_VALUE))

head(dadosSelected)
```


The dataset contains **`r nrow(dadosSelected)` monthly observations** of drinking milk production in Portugal (thousand tonnes), from January 1987 to February 2026, sourced from Eurostat (APRO_MK_COLM).

```{r}
ts_data <- ts(dadosSelected$OBS_VALUE, start = c(1987, 1), frequency = 12)
```
```{r}
ts_after<- window(ts_data, start = c(1992, 1)) #troquei em baixo no plot para se poder vizualizar
```

### 2.3 Outlier Analysis

```{r}
boxplot(dadosSelected$OBS_VALUE, main = "Boxplot of Drinking Milk Production in Portugal", ylab = "Thousand tonnes", col = "lightblue")
```

The boxplot shows no outliers. All observations fall within the expected range.

### 2.4 Time Series Visualisation

```{r}
autoplot(ts_after,
         ylab = "Thousand tonnes",
         xlab = "Year",
         main = "Monthly Drinking Milk Production in Portugal (1987-2026)")
```

From the plot we can observe:

1.  **Seasonality:** there is a clear repeating pattern every 12 months, visible throughout the whole series.
2.  **Trend**: production increases from 1987 until around 2005–2007, reaching a peak of about 90 thousand tonnes, and then steadily decreases to around 40–60 thousand tonnes by 2026.
3.  **Variance**: the seasonal swings appear to be larger in the middle of the series (around 2000–2010) than at the beginning and end, which suggests the variance may not be constant. This will be checked formally using a Box–Cox test.

### 2.5 Transformation to the month lenght

Months have different lengths (28–31 days), which introduces a spurious seasonal effect. We standardise production to an equivalent month of 30.44 days (365.25/12).

```{r}
dados_ajustados <- dadosSelected %>%
  mutate(Data_Certa = ym(TIME_PERIOD)) %>% # to make sure is in the right format
  mutate(Dias_no_Mes = days_in_month(Data_Certa)) %>%
  mutate(Adjusted_value = (OBS_VALUE* (365.25/12))/ Dias_no_Mes) 
```

```{r}
ts_adj<- ts(dados_ajustados$Adjusted_value, start = c(1987, 1), frequency = 12)
ts_adj<-window(ts_adj, start = c(1992, 1)) #se quiseres a serie toda comenta isto
```

```{r}
autoplot(ts_after, series = "Original") +
  autolayer(ts_adj , series = "Adjusted") +
  labs(
    x = "Year", 
    y = "Thousand tonnes", 
    title = "Effect of Month lenght Adjustment"
  ) +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()
```

The calendar-adjusted series removes the artefact caused by month length differences. **We will use `ts_adj` for all subsequent analysis.**

### 2.6 Data Split: Training and Test Sets

Following standard practice, we split the data into a training set (80%) for model fitting and a test set (20%) for forecast evaluation.

```{r}
n          <- length(ts_adj)
train_size <- floor(0.75 * n)

ts_train <- window(ts_adj, end   = time(ts_adj)[train_size])
ts_test  <- window(ts_adj, start = time(ts_adj)[train_size + 1])

cat("Training set:", length(ts_train), "observations\n")
cat("Test set:    ", length(ts_test),  "observations\n")
```

All transformations and model fitting are applied **only to the training set** to avoid data leakage.

### 2.7 Decomposition

```{r}
decomposed <- decompose(ts_train)
autoplot(decomposed) +
  labs(title = "Classical Decomposition of the Calendar-Adjusted Series")
```

The decomposition confirms:

-   A clear **seasonal component** with period 12 and constant amplitude throughout the series, which supports the use of an additive decomposition.
-   A **trend component**  reveals a severe structural break starting in 2015, moving from a historic plateau into a sharp, sustained decline. This structural shift, historically driven by the abolition of EU milk quotas in 2015, violates the stationarity assumptions of standard linear ARIMA models.
-   A **remainder** that fluctuates around zero with no obvious pattern, suggesting it is roughly stationary. There is a notable spike around 1989–1990, but no systematic structure is visible.

### 2.8 Variance Stabilisation

We use the Box–Cox procedure to check whether a variance-stabilising transformation is needed. A lambda close to 0 suggests a log transformation; a lambda close to 1 means no transformation is necessary.

```{r}
lambda <- BoxCox.lambda(ts_train)
cat("Optimal Box-Cox lambda:", round(lambda, 3), "\n")
```

```{r}
if (abs(lambda) < 0.25) {
  cat("Lambda close to 0 -> log transformation recommended.\n")
  ts_train_t <- log(ts_train)
  ts_test_t  <- log(ts_test)
  transformed <- TRUE
} else if (lambda > 0.75 & lambda < 1.25) {
  cat("Lambda close to 1 -> no transformation needed.\n")
  ts_train_t <- ts_train
  ts_test_t  <- ts_test
  transformed <- FALSE
} else {
  cat("Applying Box-Cox transformation with lambda =", round(lambda, 3), "\n")
  ts_train_t <- BoxCox(ts_train, lambda)
  ts_test_t  <- BoxCox(ts_test,  lambda)
  transformed <- TRUE
}

autoplot(ts_train_t,
         main = "Training Series (after transformation, if applied)",
         ylab = "Value", xlab = "Year")
```

### 2.9 Stationarity Assessment

Stationarity can be visually assessed by plotting the ACF/PACF

```{r}
tsdisplay(ts_train_t,
          main = paste0("ACF / PACF of the Series box cox transformed "),
          lag.max = 48)
```

By the results: -the sample ACF presents a periodic behavior slowly tending to zero, it is evidence of non-stationarity in seasonality and non stationary trend, which justifies the mathematical necessity of applying a regular difference ($d=1$) and a seasonal difference ($D=1$).

To confirm we use two complementary tests:

-   **ADF** (Augmented Dickey–Fuller): H₀ = series has a unit root (non-stationary).
-   **KPSS**: H₀ = series is stationary.

```{r}
adf.test(ts_train_t)
kpss.test(ts_train_t)
```

```{r}
cat("Recommended regular differences (ndiffs):  ", ndiffs(ts_train_t),  "\n")
cat("Recommended seasonal differences (nsdiffs):", nsdiffs(ts_train_t), "\n")
```

Both tests agree that the series is non-stationary:

-   **ADF** (p-value = 0.385): fails to reject H₀, indicating the presence of a unit root.
-   **KPSS** (p-value \< 0.01): rejects H₀, confirming non-stationarity.

The functions `ndiffs()` and `nsdiffs()` recommend **d = 1** regular difference and **D = 1** seasonal difference to achieve stationarity.

### 2.10 Achieving Stationarity through Differencing

We first apply a seasonal difference (lag = 12) to remove the seasonal pattern, then a regular first difference to remove the trend.

```{r}
d  <- ndiffs(ts_train_t)
D  <- nsdiffs(ts_train_t)

ts_stationary <- ts_train_t

if (D > 0) {
  ts_stationary <- diff(ts_stationary, lag = 12, differences = D)
}
if (d > 0) {
  ts_stationary <- diff(ts_stationary, differences = d)
}

autoplot(ts_stationary,
         main = paste0("Stationary Series (d=", d, ", D=", D, ")"),
         ylab = "Value", xlab = "Year")
```

We re-run the stationarity tests to confirm.

```{r}
adf.test(ts_stationary)
kpss.test(ts_stationary)
```

After applying one seasonal difference (D=1, lag=12) and one regular difference (d=1), both tests confirm that the series is now stationary:

-   **ADF** (p-value \< 0.01): rejects H₀, confirming stationarity.
-   **KPSS** (p-value \> 0.1): fails to reject H₀, confirming stationarity.

The plot shows the series fluctuating around zero with no visible trend or seasonality. Some larger spikes are visible in the early years (1988–1993), but no systematic pattern remains.

```{r}
ts_stationary
```

### 2.11 ACF and PACF Analysis

The ACF and PACF of the stationary series are used to identify candidate SARIMA orders.

```{r}
tsdisplay(ts_stationary,
          main = paste0("ACF / PACF of Stationary Series (d=", d, ", D=", D, ")"),
          lag.max = 48)
```

For a seasonal series with s = 12, the relevant lags to examine are **lag 1** (non-seasonal structure) and **lags 12 and 24** (seasonal structure).
Non seasonal order:
By looking at the graphs none show either a strong decay or a cut off, p=1, q=1
Seasonal order:
In the seasonal lags both show negative significantly spikes in lag 12 which  suggests a seasonal MA(1) model

`auto.arima()` will also be used in Section 3 to confirm or refine this choice.

------------------------------------------------------------------------

## 3. Model Proposals

All models are fitted on the log-transformed training set (`ts_train_t`). The transformation will be reversed when forecasting in Section 4.

### 3.1 Auto ARIMA

We first use `auto.arima()` to search broadly over SARIMA orders and find the best model according to AICc. This serves as a reference to compare against the manually identified model.

Autoarima

```{r}
auto_fit <- auto.arima(ts_train_t, stepwise = FALSE, approximation = FALSE)
summary(auto_fit)
```

The selected model is **SARIMA(1,1,1)(0,1,1)[12]**. This is consistent with the ACF/PACF analysis, the presence of spikes at lag 1 in both plots suggested a mixed ARMA(1,1) non-seasonal structure rather than a pure MA(1). All three coefficients are estimated with low standard errors, suggesting they are well-determined by the data. The training MAPE of 0.92% indicates an excellent in-sample fit. **Com o que ta escrito aqui ver em cima**

#### Parameter significance and correlation

```{r}
matriz_vcov <- vcov(auto_fit)
matriz_correlacao <- cov2cor(matriz_vcov)

print(matriz_correlacao)
```

The absolute correlation between any two parameters must be less than $0.70$.

The highest correlation observed in the model's correlation matrix was $-0.68$. Since $|-0.68| = 0.68 < 0.70$, the model satisfies this empirical rule. While close to the threshold, this result indicates no strong evidence of overparameterization or redundancy.

```{r}
coeftest(auto_fit)
```

#### Residual diagnostics

```{r}
checkresiduals(auto_fit)
```

All three parameters are highly significant (p \< 0.001). The residual plot shows values fluctuating around zero with no obvious trend or seasonality. The histogram is approximately bell-shaped, consistent with normality. However, the **Ljung-Box test** returns a p-value of **0.03**, which is below the 0.05 threshold, meaning we reject the null hypothesis of no autocorrelation. Two borderline spikes are present but do not follow a systematic seasonal pattern, suggesting any remaining autocorrelation is weak. This result motivates testing alternative models.

#### Forecast Auto Arima

```{r}
forecast_horizon <- length(ts_test_t)
sarima_forecast <- forecast(auto_fit, h = forecast_horizon)

autoplot(sarima_forecast, include = forecast_horizon) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "(Auto.Arima) ARIMA(1,1,1)(0,1,1)[12] Forecast vs. Actual Data",
       x = "Year", y = "Production Thousand tonnes") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()
```




```{r}
accuracy(sarima_forecast, ts_test_t)
```
low values of MPE and MAPE  in test ate 10 é bom , negative values in mpe means is predicting lower than reality

 


### 3.2 Manual SARIMA — SARIMA(1,1,1)(1,1,1)[12] 


```{r}
manual_fit <- Arima(ts_train_t,
                    order    = c(1, 1, 1),
                    seasonal = c(1, 1, 1),
                    method = "ML"
                    )
summary(manual_fit)
```

```{r}
forecast::checkresiduals(manual_fit)

cov2cor(vcov(manual_fit))
```

#### Parameter significance

```{r}
coeftest(manual_fit)
```

#### Residual diagnostics

```{r}
checkresiduals(manual_fit)
```

Both coefficients are highly significant (p \< 2.2e-16). The residuals fluctuate around zero with no visible trend or seasonal pattern. However, the **Ljung-Box test** returns a significant p-value (p \< 0.05), indicating residual autocorrelation, which means that the model leaves some structure unexplained. The ACF of the residuals shows six spikes exceeding the confidence bands, suggesting that the airline model is not fully capturing the dynamics of the series - it underfits the data, being too simple for the complexity of this series. This motivates exploring models with AR terms in the next section.

#### Forecast 
```{r}
forecast_horizon <- length(ts_test_t)
sarima_manual_forecast <- forecast(manual_fit, h = forecast_horizon)

autoplot(sarima_forecast, include = forecast_horizon) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "ARIMA(1,1,1)(1,1,1)[12] Forecast vs. Actual Data",
       x = "Year", y = "Production Thousand tonnes") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()
```

```{r}
accuracy(sarima_manual_forecast, ts_test_t)
```

### 3.3 Candidate Models with AR Terms

To address the residual autocorrelation found in the airline model, we add non-seasonal AR terms. Three candidates are fitted and compared against SARIMA(0,1,1)(0,1,1)[12].

```{r}
fit_111_011 <- Arima(ts_train_t, order = c(1,1,1), seasonal = c(0,1,1))
fit_211_011 <- Arima(ts_train_t, order = c(2,1,1), seasonal = c(0,1,1))
fit_112_011 <- Arima(ts_train_t, order = c(1,1,2), seasonal = c(0,1,1))
```

#### AIC comparison

```{r}
ar_candidates <- data.frame(
  Model = c(
    "SARIMA(0,1,1)(0,1,1)[12]",
    "SARIMA(1,1,1)(0,1,1)[12]",
    "SARIMA(2,1,1)(0,1,1)[12]",
    "SARIMA(1,1,2)(0,1,1)[12]"
  ),
  AIC  = round(c(AIC(manual_fit), AIC(fit_111_011), AIC(fit_211_011), AIC(fit_112_011)), 2),
  AICc = round(c(AICc(manual_fit), AICc(fit_111_011), AICc(fit_211_011), AICc(fit_112_011)), 2),
  BIC  = round(c(BIC(manual_fit), BIC(fit_111_011), BIC(fit_211_011), BIC(fit_112_011)), 2)
)

ar_candidates$ΔAIC <- round(ar_candidates$AIC - min(ar_candidates$AIC), 2)
ar_candidates[order(ar_candidates$AIC), ] %>%
  kable(row.names = FALSE, caption = "AIC Comparison — SARIMA Candidates with AR Terms") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

SARIMA(1,1,1)(0,1,1)[12] achieves the lowest AIC among all non-seasonal AR candidates — consistent with the model selected by `auto.arima()` in Section 3.1. However, the Ljung-Box test still returns p = 0.03, motivating two further adjustments in the next section.

### 3.4 Extended Seasonal Candidates

We test two further adjustments targeting the remaining residual autocorrelation: adding a seasonal AR term (P=1) and adding a second seasonal MA term (Q=2).

```{r}
fit_111_111 <- Arima(ts_train_t, order = c(1,1,1), seasonal = list(order = c(1,1,1), period = 12))
fit_111_012 <- Arima(ts_train_t, order = c(1,1,1), seasonal = list(order = c(0,1,2), period = 12))
```

#### Residual diagnostics — SARIMA(1,1,1)(1,1,1)[12]

```{r}
checkresiduals(fit_111_111)
```

#### Parameter significance — SARIMA(1,1,1)(1,1,1)[12]

```{r}
coeftest(fit_111_111)
```

The residuals fluctuate around zero with no visible trend or seasonal pattern, and the histogram is approximately normal. However, two spikes still exceed the confidence bands in the ACF, and the **Ljung-Box test** returns p = 0.021, meaning that the residual autocorrelation is not resolved. Furthermore, the seasonal AR term `sar1` is not statistically significant (p = 0.947), with an estimate virtually equal to zero, confirming that this parameter adds no meaningful information to the model.

#### Residual diagnostics — SARIMA(1,1,1)(0,1,2)[12]

```{r}
checkresiduals(fit_111_012)
```

#### Parameter significance — SARIMA(1,1,1)(0,1,2)[12]

```{r}
coeftest(fit_111_012)
```

The residuals fluctuate around zero with no visible trend or seasonal pattern, and the histogram is approximately normal. Again, two spikes still exceed the confidence bands in the ACF, and the **Ljung-Box test** returns p = 0.021 which is identical to the previous candidate. The second seasonal MA term `sma2` is not statistically significant (p = 0.946), with an estimate virtually equal to zero. The additional parameter does not capture any real structure in the data.

#### AIC comparison — extended candidates

```{r}
extended_candidates <- data.frame(
  Model = c(
    "SARIMA(0,1,1)(0,1,1)[12]",
    "SARIMA(1,1,1)(0,1,1)[12]",
    "SARIMA(2,1,1)(0,1,1)[12]",
    "SARIMA(1,1,2)(0,1,1)[12]",
    "SARIMA(1,1,1)(1,1,1)[12]",
    "SARIMA(1,1,1)(0,1,2)[12]"
  ),
  AIC  = round(c(AIC(manual_fit), AIC(fit_111_011), AIC(fit_211_011), AIC(fit_112_011),
                 AIC(fit_111_111), AIC(fit_111_012)), 2),
  AICc = round(c(AICc(manual_fit), AICc(fit_111_011), AICc(fit_211_011), AICc(fit_112_011),
                 AICc(fit_111_111), AICc(fit_111_012)), 2),
  BIC  = round(c(BIC(manual_fit), BIC(fit_111_011), BIC(fit_211_011), BIC(fit_112_011),
                 BIC(fit_111_111), BIC(fit_111_012)), 2)
)

extended_candidates$ΔAIC <- round(extended_candidates$AIC - min(extended_candidates$AIC), 2)
extended_candidates[order(extended_candidates$AIC), ] %>%
  kable(row.names = FALSE, caption = "AIC Comparison — All SARIMA Candidates") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**SARIMA(1,1,1)(0,1,1)[12]** achieves the lowest AIC across all candidates (ΔAIC = 0). The additional seasonal parameters in SARIMA(1,1,1)(1,1,1)[12] and SARIMA(1,1,1)(0,1,2)[12] are not statistically significant and do not improve either the AIC or the Ljung-Box test, confirming that the extra complexity is not justified. This model is therefore selected as the final SARIMA model.


### 3.5 Structural Break Diagnostic — Pre-2015 Model

The decomposition revealed a severe structural break starting in 2015, coinciding with the abolition of EU milk quotas. To assess whether the SARIMA model is adequate for the stable pre-break period, we fit the same SARIMA(1,1,1)(0,1,1)[12] on training data restricted to before 2015. If the model performs significantly better on this subset, it confirms that the structural break — not the model specification — is responsible for the residual autocorrelation that persisted across all candidate models.

```{r}
ts_train_pre2015 <- window(ts_train_t, end = c(2014, 12))

fit_pre2015 <- Arima(ts_train_pre2015,
                     order    = c(1, 1, 1),
                     seasonal = c(0, 1, 1))
summary(fit_pre2015)
```

### 

```{r}
cov2cor(vcov(fit_pre2015))
```

#### Parameter significance — Pre-2015 model

```{r}
coeftest(fit_pre2015)
```

#### Residual diagnostics — Pre-2015 model

```{r}
checkresiduals(fit_pre2015)
```

The pre-2015 model yields a marginal improvement in the Ljung-Box test (p = 0.049 vs p = 0.03 for the full series), just crossing the 0.05 threshold. While this provides some evidence that the 2015 structural break contributes to the residual autocorrelation, the improvement is not dramatic. The notable volatility in the residuals during 1988–1992 suggests that data quality in the early years may also be a contributing factor. Nonetheless, the result supports the decision to explicitly model the structural break in Section 3.6.

---

### 3.6 ARIMAX — SARIMA(1,1,1)(0,1,1)[12] with Structural Break Dummy

To model the full series while accounting for the 2015 structural break, we extend SARIMA(1,1,1)(0,1,1)[12] with an exogenous step-change dummy variable (0 before 2015, 1 from January 2015 onwards). This formulation, known as ARIMAX, estimates the permanent level shift caused by the abolition of EU milk quotas as a regression coefficient, allowing the ARIMA errors to remain stationary.

```{r}
dummy_train <- as.numeric(time(ts_train_t) >= 2015)
dummy_test  <- as.numeric(time(ts_test_t)  >= 2015)

fit_arimax <- Arima(ts_train_t,
                    order    = c(1, 1, 1),
                    seasonal = c(0, 1, 1),
                    xreg     = dummy_train)
summary(fit_arimax)
```

### 

```{r}
cov2cor(vcov(fit_arimax))
```

#### Parameter significance — ARIMAX

```{r}
coeftest(fit_arimax)
```

#### Residual diagnostics — ARIMAX

```{r}
checkresiduals(fit_arimax)
```

#### AIC comparison — SARIMA vs ARIMAX

```{r}
break_comparison <- data.frame(
  Model = c("SARIMA(1,1,1)(0,1,1)[12]", "ARIMAX(1,1,1)(0,1,1)[12] + break dummy"),
  AIC   = round(c(AIC(fit_111_011), AIC(fit_arimax)), 2),
  AICc  = round(c(AICc(fit_111_011), AICc(fit_arimax)), 2),
  BIC   = round(c(BIC(fit_111_011), BIC(fit_arimax)), 2)
)
break_comparison$ΔAIC <- round(break_comparison$AIC - min(break_comparison$AIC), 2)
break_comparison %>%
  kable(row.names = FALSE, caption = "SARIMA vs ARIMAX with Structural Break Dummy") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

#### Forecast — ARIMAX

```{r}
arimax_forecast <- forecast(fit_arimax,
                             h    = length(ts_test_t),
                             xreg = dummy_test)

autoplot(arimax_forecast, include = length(ts_test_t)) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "ARIMAX(1,1,1)(0,1,1)[12] with Structural Break — Forecast vs. Actual",
       x = "Year", y = "Value (log-transformed)") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()
```

```{r}
accuracy(arimax_forecast, ts_test_t)
```

#### Ramp Dummy Variant

To address this limitation, we replace the step dummy with a **ramp variable** that grows linearly from zero after January 2015. Unlike a step dummy, the ramp captures a sustained, continuously changing trend rather than a one-time level shift.

```{r}
time_idx   <- as.numeric(time(ts_train_t))
ramp_train <- pmax(0, time_idx - 2015)

time_idx_test <- as.numeric(time(ts_test_t))
ramp_test     <- pmax(0, time_idx_test - 2015)

fit_arimax_ramp <- Arima(ts_train_t,
                         order    = c(1, 1, 1),
                         seasonal = c(0, 1, 1),
                         xreg     = ramp_train)
summary(fit_arimax_ramp)
```
### 

```{r}
cov2cor(vcov(fit_arimax))
```

#### Parameter significance — Ramp ARIMAX

```{r}
coeftest(fit_arimax_ramp)
```

#### Residual diagnostics — Ramp ARIMAX

```{r}
checkresiduals(fit_arimax_ramp)
```

#### AIC comparison — Step vs Ramp

```{r}
ramp_comparison <- data.frame(
  Model = c(
    "SARIMA(1,1,1)(0,1,1)[12]",
    "ARIMAX — step dummy",
    "ARIMAX — ramp dummy"
  ),
  AIC  = round(c(AIC(fit_111_011), AIC(fit_arimax), AIC(fit_arimax_ramp)), 2),
  AICc = round(c(AICc(fit_111_011), AICc(fit_arimax), AICc(fit_arimax_ramp)), 2),
  BIC  = round(c(BIC(fit_111_011), BIC(fit_arimax), BIC(fit_arimax_ramp)), 2)
)
ramp_comparison$ΔAIC <- round(ramp_comparison$AIC - min(ramp_comparison$AIC), 2)
ramp_comparison %>%
  kable(row.names = FALSE, caption = "AIC Comparison — SARIMA vs ARIMAX Variants") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

#### Forecast — Ramp ARIMAX

```{r}
arimax_ramp_forecast <- forecast(fit_arimax_ramp,
                                  h    = length(ts_test_t),
                                  xreg = ramp_test)

autoplot(arimax_ramp_forecast, include = length(ts_test_t)) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "ARIMAX(1,1,1)(0,1,1)[12] with Ramp Dummy — Forecast vs. Actual",
       x = "Year", y = "Value (log-transformed)") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()
```

```{r}
accuracy(arimax_ramp_forecast, ts_test_t)
```

The ramp dummy coefficient is not statistically significant (xreg = −0.003, p = 0.930), and the AIC deteriorates relative to both the plain SARIMA and the step dummy ARIMAX. This failure has a structural explanation: the model applies d = 1 and D = 1 differencing to the error process. After one regular difference, a linear ramp reduces to a constant; after one further seasonal difference, that constant is eliminated. The differencing therefore neutralises the ramp's intended effect, leaving it with no explanatory power.

Comparing the three models on the test set, neither ARIMAX variant achieves a Theil's U below 1 — both perform worse than a naïve forecast out-of-sample. The step dummy is preferred on in-sample grounds (lowest AIC, significant coefficient), while the ramp dummy offers a marginally lower test MAPE (1.78% vs 2.62%) despite its insignificant coefficient, an artefact of the declining forecast trajectory rather than genuine model improvement. The fundamental limitation is that the 2015 structural break introduced a persistent change in the trend of the series that ARIMA-based models, which assume stationarity after differencing, cannot fully accommodate. This motivates comparing the ARIMAX against the ETS model in Section 3.7, which adapts its level and trend components dynamically and may handle this structural shift more naturally.

---

### 3.7 ETS Model

We will try another model, starting with ets auto
```{r}
ets_fit <- ets(ts_train_t,lambda=0) #making sure it doesn't apply another box cox transformation

summary(ets_fit)
```
Addictive Holt-Winters method 



```{r}
checkresiduals(ets_fit)
```
very bad in the Ljung-Box test, need the diferation


```{r}
forecast_horizon <- length(ts_test_t)
ets_forecast <- forecast(ets_fit, h = forecast_horizon)

autoplot(ets_forecast, include = forecast_horizon) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "ETS(A,A,A)  Forecast vs. Actual Data",
       x = "Year", y = "Production Thousand tonnes") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()


```
```{r}
accuracy(ets_forecast, ts_test_t)
```


### 3.8 STLM mix of both (etls and arima)

```{r}
stlm_arima_fit <- stlm(ts_train_t, s.window = "periodic", method = "arima")


checkresiduals(stlm_arima_fit)

# 3. Ver que ordens de ARIMA foram usadas internamente para corrigir a série
stlm_arima_fit$model
```
Still fails the Ljung-Box test but way improved, does worse than sarima


```{r}
forecast_horizon <- length(ts_test_t)
stlm_forecast <- forecast(stlm_arima_fit, h = forecast_horizon)

autoplot(stlm_forecast, include = forecast_horizon) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "STLM_Arima Forecast vs. Actual Data",
       x = "Year", y = "Production Thousand tonnes") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()

accuracy(stlm_forecast,ts_test_t)

```
Does worse in forecast 




### TBATS

```{r}
fit_tbats <- tbats(ts_train_t)
summary(fit_tbats)
tbats_forecast <- forecast(fit_tbats, h = length(ts_test_t))

autoplot(tbats_forecast, include = length(ts_test_t)) +
  autolayer(ts_test_t, series = "Actual Data") +
  labs(title = "TBATS Forecast vs. Actual Data", x = "Year", y = "Value") +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()

accuracy(tbats_forecast, ts_test_t)
checkresiduals(fit_tbats)
```

## Comparision
```{r}
criteria <- data.frame(
  Model = c("SARIMA_auto", "SARIMA_manual", "ETS", "STLM_ARIMA"),
  AIC = c(AIC(auto_fit), AIC(manual_fit), ets_fit$aic, AIC(stlm_arima_fit$model)),
  AICc = c(AICc(auto_fit), AICc(manual_fit), ets_fit$aicc, AICc(stlm_arima_fit$model)),
  BIC = c(BIC(auto_fit), BIC(manual_fit), ets_fit$bic, BIC(stlm_arima_fit$model))
)

criteria_num <- criteria
criteria_num[, -1] <- round(criteria[, -1], 2)
kable(criteria_num, caption = "Comparison of Information Criteria across Models") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

```{r}
library(knitr)

results <- rbind(
  SARIMA_auto = accuracy(sarima_forecast, ts_test_t)[2, c("RMSE", "MAE", "MAPE", "Theil's U")],
  SARIMA_manual = accuracy(sarima_manual_forecast, ts_test_t)[2, c("RMSE", "MAE", "MAPE", "Theil's U")],
  ETS = accuracy(ets_forecast, ts_test_t)[2, c("RMSE", "MAE", "MAPE", "Theil's U")],
  STLM_ARIMA = accuracy(stlm_forecast, ts_test_t)[2, c("RMSE", "MAE", "MAPE", "Theil's U")]
)
kable(round(results, 2), caption = "Forecast Accuracy Comparison on Test Set") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
