---
title: "Time Series Analysis of Monthly Drinking Milk Production in Portugal"
output: html_notebook
---

## 1. Introduction

*(to be written)*

---

## 2. Data and Exploratory Analysis

### 2.1 Packages

Necessary packages
```{r message=FALSE, warning=FALSE}
#install.packages(c("readxl","lubridate","dplyr","ggplot2","forecast","tseries","urca","lmtest","MuMIn","kableExtra"))

library(readxl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(urca)
library(lmtest)
library(MuMIn)
library(kableExtra)
```

### 2.2 Data Loading and Preparation

```{r}
dados <- read.csv("data/leite.csv")
dados
```

Select the relevant columns and remove the three empty rows at the start (1986, no data reported).

```{r}
dadosSelected <- select(dados, TIME_PERIOD, OBS_VALUE)
dadosSelected <- dadosSelected[-c(1,2, 3), ]

# Ensure numeric values (handles any locale comma decimals)
dadosSelected$OBS_VALUE <- as.numeric(gsub(",", ".", dadosSelected$OBS_VALUE))

head(dadosSelected)
```

The dataset contains **`r nrow(dadosSelected)` monthly observations** of drinking milk production in Portugal (thousand tonnes), from January 1987 to February 2026, sourced from Eurostat (APRO_MK_COLM).

```{r}
ts_data <- ts(dadosSelected$OBS_VALUE, start = c(1987, 1), frequency = 12)
```

### 2.3 Outlier Analysis

```{r}
boxplot(dadosSelected$OBS_VALUE, main = "Boxplot of Drinking Milk Production in Portugal", ylab = "Thousand tonnes", col = "lightblue")
```

The boxplot shows no outliers. All observations fall within the expected range.

### 2.4 Time Series Visualisation

```{r}
autoplot(ts_data,
         ylab = "Thousand tonnes",
         xlab = "Year",
         main = "Monthly Drinking Milk Production in Portugal (1987-2026)")
```
From the plot we can observe:

1. **Seasonality:** there is a clear repeating pattern every 12 months, visible throughout the whole series.
2. **Trend**: production increases from 1987 until around 2005–2007, reaching a peak of about 90 thousand tonnes, and then steadily decreases to around 40–60 thousand tonnes by 2026.                          
3. **Variance**: the seasonal swings appear larger in the middle of the series (around 2000–2010) than at the beginning and end, which suggests the variance may not be constant. This will be checked formally using a Box–Cox test. 

### 2.5 Transformation to the month lenght 

Months have different lengths (28–31 days), which introduces a spurious seasonal effect. We standardise production to an equivalent month of 30.44 days (365.25/12).

```{r}
dados_ajustados <- dadosSelected %>%
  # 1. Garante que o TIME_PERIOD seja lido como data (adicionando o dia 01)
  mutate(Data_Certa = ym(TIME_PERIOD)) %>%
  # 2. Calcula o número de dias que aquele mês específico teve
  mutate(Dias_no_Mes = days_in_month(Data_Certa)) %>%
  # 3. Faz o ajuste: valor original dividido pelo número de dias
  mutate(Valor_Por_Dia = (OBS_VALUE* (365.25/12))/ Dias_no_Mes)
```

```{r}
# Criamos a série usando a nova coluna 'Valor_Por_Dia'
ts_adj<- ts(dados_ajustados$Valor_Por_Dia, start = c(1987, 1), frequency = 12)
```

```{r}
# Desenha o gráfico comparativo
autoplot(ts_data, series = "Original") +
  autolayer(ts_adj , series = "Calendar-adjusted") +
  labs(
    x = "Year", 
    y = "Thousand tonnes", 
    title = "Effect of Calendar Adjustment"
  ) +
  guides(colour = guide_legend(title = "Series")) +
  theme_minimal()
```

The calendar-adjusted series removes the artefact caused by month length differences. **We will use `ts_adj` for all subsequent analysis.**

### 2.6 Data Split: Training and Test Sets

Following standard practice, we split the data into a training set (80%) for model fitting and a test set (20%) for forecast evaluation.

```{r}
n          <- length(ts_adj)
train_size <- floor(0.8 * n)

ts_train <- window(ts_adj, end   = time(ts_adj)[train_size])
ts_test  <- window(ts_adj, start = time(ts_adj)[train_size + 1])

cat("Training set:", length(ts_train), "observations\n")
cat("Test set:    ", length(ts_test),  "observations\n")
```

All transformations and model fitting are applied **only to the training set** to avoid data leakage.

### 2.7 Decomposition


```{r}
decomposed <- decompose(ts_train)
autoplot(decomposed) +
  labs(title = "Classical Decomposition of the Calendar-Adjusted Series")
```

The decomposition confirms:

- A clear **seasonal component** with period 12 and constant amplitude throughout the series, which supports the use of an additive decomposition.
- A **trend component** that rises steadily from 1987 to a peak around 2005–2007 (~80 thousand tonnes), then declines consistently through to the end of the training set.
- A **remainder** that fluctuates around zero with no obvious pattern, suggesting it is roughly stationary. There is a notable spike around 1989–1990, but no systematic structure is visible.

### 2.8 Variance Stabilisation

We use the Box–Cox procedure to check whether a variance-stabilising transformation is needed. A lambda close to 0 suggests a log transformation; a lambda close to 1 means no transformation is necessary.

```{r}
lambda <- BoxCox.lambda(ts_train)
cat("Optimal Box-Cox lambda:", round(lambda, 3), "\n")
```

```{r}
if (abs(lambda) < 0.25) {
  cat("Lambda close to 0 -> log transformation recommended.\n")
  ts_train_t <- log(ts_train)
  ts_test_t  <- log(ts_test)
  transformed <- TRUE
} else if (lambda > 0.75 & lambda < 1.25) {
  cat("Lambda close to 1 -> no transformation needed.\n")
  ts_train_t <- ts_train
  ts_test_t  <- ts_test
  transformed <- FALSE
} else {
  cat("Applying Box-Cox transformation with lambda =", round(lambda, 3), "\n")
  ts_train_t <- BoxCox(ts_train, lambda)
  ts_test_t  <- BoxCox(ts_test,  lambda)
  transformed <- TRUE
}

autoplot(ts_train_t,
         main = "Training Series (after transformation, if applied)",
         ylab = "Value", xlab = "Year")
```

### 2.9 Stationarity Assessment

We use two complementary tests:

- **ADF** (Augmented Dickey–Fuller): H₀ = series has a unit root (non-stationary).
- **KPSS**: H₀ = series is stationary.

```{r}
adf.test(ts_train_t)
kpss.test(ts_train_t)
```

```{r}
cat("Recommended regular differences (ndiffs):  ", ndiffs(ts_train_t),  "\n")
cat("Recommended seasonal differences (nsdiffs):", nsdiffs(ts_train_t), "\n")
```

Both tests agree that the series is non-stationary:

- **ADF** (p-value = 0.385): fails to reject H₀, indicating the presence of a unit root.
- **KPSS** (p-value < 0.01): rejects H₀, confirming non-stationarity.

The functions `ndiffs()` and `nsdiffs()` recommend **d = 1** regular difference and **D = 1** seasonal difference to achieve stationarity.

### 2.10 Achieving Stationarity through Differencing

We first apply a seasonal difference (lag = 12) to remove the seasonal pattern, then a regular first difference to remove the trend.

```{r}
d  <- ndiffs(ts_train_t)
D  <- nsdiffs(ts_train_t)

ts_stationary <- ts_train_t

if (D > 0) {
  ts_stationary <- diff(ts_stationary, lag = 12, differences = D)
}
if (d > 0) {
  ts_stationary <- diff(ts_stationary, differences = d)
}

autoplot(ts_stationary,
         main = paste0("Stationary Series (d=", d, ", D=", D, ")"),
         ylab = "Value", xlab = "Year")
```

We re-run the stationarity tests to confirm.

```{r}
adf.test(ts_stationary)
kpss.test(ts_stationary)
```

After applying one seasonal difference (D=1, lag=12) and one regular difference (d=1), both tests confirm that the series is now stationary:

- **ADF** (p-value < 0.01): rejects H₀, confirming stationarity.
- **KPSS** (p-value > 0.1): fails to reject H₀, confirming stationarity.

The plot shows the series fluctuating around zero with no visible trend or seasonality. Some larger spikes are visible in the early years (1988–1993), but no systematic pattern remains.

### 2.11 ACF and PACF Analysis

The ACF and PACF of the stationary series are used to identify candidate SARIMA orders.

```{r}
tsdisplay(ts_stationary,
          main = paste0("ACF / PACF of Stationary Series (d=", d, ", D=", D, ")"),
          lag.max = 48)
```

For a seasonal series with s = 12, the relevant lags to examine are **lag 1** (non-seasonal structure) and **lags 12 and 24** (seasonal structure).

- **Lag 1**: both the ACF and PACF show a single significant negative spike at lag 1, with lags 2–10 falling within the confidence bands. This is consistent with a non-seasonal **MA(1)** component (q = 1).
- **Lag 12**: both the ACF and PACF show a significant negative spike at lag 12, indicating a seasonal **MA(1)** component (Q = 1).
- **Lag 24**: no significant spikes in either ACF or PACF, confirming that one seasonal MA term is sufficient (P = 0).

The pattern — one spike at lag 1 and one spike at lag 12, with nothing beyond — directly suggests the **airline model**:

- **SARIMA(0,1,1)(0,1,1)[12]**

`auto.arima()` will also be used in Section 3 to confirm or refine this choice.

---
## 3. Model Proposals

All models are fitted on the log-transformed training set (`ts_train_t`). The transformation will be reversed when forecasting in Section 4.

### 3.1 Auto ARIMA

We first use `auto.arima()` to search broadly over SARIMA orders and find the best model according to AICc. This serves as a reference to compare against the manually identified model.

```{r}
auto_fit <- auto.arima(ts_train_t, stepwise = FALSE, approximation = FALSE)
summary(auto_fit)
```

The selected model is **SARIMA(1,1,1)(0,1,1)[12]**. This is consistent with the ACF/PACF analysis — the presence of spikes at lag 1 in both plots suggested a mixed ARMA(1,1) non-seasonal structure rather than a pure MA(1). All three coefficients are estimated with low standard errors, suggesting they are well-determined by the data. The training MAPE of 0.92% indicates an excellent in-sample fit.

#### Parameter significance

```{r}
coeftest(auto_fit)
```

#### Residual diagnostics

```{r}
checkresiduals(auto_fit)
```
All three parameters are highly significant (p < 0.001). The residual plot shows values fluctuating around zero with no obvious trend or seasonality. The histogram is approximately bell-shaped, consistent with normality. However, the **Ljung-Box test** returns a p-value of **0.03**, which is below the 0.05 threshold, meaning we reject the null hypothesis of no autocorrelation. Two borderline spikes are present but do not follow a systematic seasonal pattern, suggesting any remaining autocorrelation is weak. This result motivates testing alternative models. 

---

### 3.2 Manual SARIMA — SARIMA(0,1,1)(0,1,1)[12]

Based on the ACF/PACF analysis in Section 2, the pattern of a single significant negative spike at lag 1 and at lag 12 suggesting SARIMA(0,1,1)(0,1,1)[12].

```{r}
manual_fit <- Arima(ts_train_t,
                    order    = c(0, 1, 1),
                    seasonal = list(order = c(0, 1, 1), period = 12))
summary(manual_fit)
```
#### Parameter significance

```{r}
coeftest(manual_fit)
```

#### Residual diagnostics

```{r}
checkresiduals(manual_fit)
```
Both coefficients are highly significant (p < 2.2e-16). The residuals fluctuate around zero with no visible trend or seasonal pattern. However, the **Ljung-Box test** returns a significant p-value (p < 0.05), indicating residual autocorrelation, which means that the model leaves some structure unexplained. The ACF of the residuals shows six spikes exceeding the confidence bands, suggesting that the airline model is not fully capturing the dynamics of the series — it underfits the data, being too simple for the complexity of this series. This motivates exploring models with AR terms in the next section.

---

### 3.3 Candidate Models with AR Terms                                                                                                                                                                         
To address the residual autocorrelation found in the airline model, we add non-seasonal AR terms. Three candidates are fitted and compared against SARIMA(0,1,1)(0,1,1)[12].

```{r}
fit_111_011 <- Arima(ts_train_t, order = c(1,1,1), seasonal = c(0,1,1))
fit_211_011 <- Arima(ts_train_t, order = c(2,1,1), seasonal = c(0,1,1))
fit_112_011 <- Arima(ts_train_t, order = c(1,1,2), seasonal = c(0,1,1))
```

#### AIC comparison

```{r}
ar_candidates <- data.frame(
  Model = c(
    "SARIMA(0,1,1)(0,1,1)[12]",
    "SARIMA(1,1,1)(0,1,1)[12]",
    "SARIMA(2,1,1)(0,1,1)[12]",
    "SARIMA(1,1,2)(0,1,1)[12]"
  ),
  AIC  = round(c(AIC(manual_fit), AIC(fit_111_011), AIC(fit_211_011), AIC(fit_112_011)), 2),
  AICc = round(c(AICc(manual_fit), AICc(fit_111_011), AICc(fit_211_011), AICc(fit_112_011)), 2),
  BIC  = round(c(BIC(manual_fit), BIC(fit_111_011), BIC(fit_211_011), BIC(fit_112_011)), 2)
)

ar_candidates$ΔAIC <- round(ar_candidates$AIC - min(ar_candidates$AIC), 2)
ar_candidates[order(ar_candidates$AIC), ] %>%
  kable(row.names = FALSE, caption = "AIC Comparison — SARIMA Candidates with AR Terms") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

SARIMA(1,1,1)(0,1,1)[12] achieves the lowest AIC among all non-seasonal AR candidates — consistent with the model selected by `auto.arima()` in Section 3.1. However, the Ljung-Box test still returns p = 0.03, motivating two further adjustments in the next section.

---

### 3.4 Extended Seasonal Candidates

We test two further adjustments targeting the remaining residual autocorrelation: adding a seasonal AR term (P=1) and adding a second seasonal MA term (Q=2).

```{r}
fit_111_111 <- Arima(ts_train_t, order = c(1,1,1), seasonal = list(order = c(1,1,1), period = 12))
fit_111_012 <- Arima(ts_train_t, order = c(1,1,1), seasonal = list(order = c(0,1,2), period = 12))
```

#### Residual diagnostics — SARIMA(1,1,1)(1,1,1)[12]

```{r}
checkresiduals(fit_111_111)
```

#### Parameter significance — SARIMA(1,1,1)(1,1,1)[12]

```{r}
coeftest(fit_111_111)
```
The residuals fluctuate around zero with no visible trend or seasonal pattern, and the histogram is approximately normal. However, two spikes still exceed the confidence bands in the ACF, and the **Ljung-Box test** returns p = 0.021, meaning that the residual autocorrelation is not resolved. Furthermore, the seasonal AR term `sar1` is not statistically significant (p = 0.947), with an estimate virtually equal to zero, confirming that this parameter adds no meaningful information to the model.

#### Residual diagnostics — SARIMA(1,1,1)(0,1,2)[12]

```{r}
checkresiduals(fit_111_012)
```

#### Parameter significance — SARIMA(1,1,1)(0,1,2)[12]

```{r}
coeftest(fit_111_012)
```
The residuals fluctuate around zero with no visible trend or seasonal pattern, and the histogram is approximately normal. Again, two spikes still exceed the confidence bands in the ACF, and the **Ljung-Box test** returns p = 0.021 which is identical to the previous candidate. The second seasonal MA term `sma2` is not statistically significant (p = 0.946), with an estimate virtually equal to zero. The additional parameter does not capture any real structure in the data.

#### AIC comparison — extended candidates

```{r}
extended_candidates <- data.frame(
  Model = c(
    "SARIMA(0,1,1)(0,1,1)[12]",
    "SARIMA(1,1,1)(0,1,1)[12]",
    "SARIMA(2,1,1)(0,1,1)[12]",
    "SARIMA(1,1,2)(0,1,1)[12]",
    "SARIMA(1,1,1)(1,1,1)[12]",
    "SARIMA(1,1,1)(0,1,2)[12]"
  ),
  AIC  = round(c(AIC(manual_fit), AIC(fit_111_011), AIC(fit_211_011), AIC(fit_112_011),
                 AIC(fit_111_111), AIC(fit_111_012)), 2),
  AICc = round(c(AICc(manual_fit), AICc(fit_111_011), AICc(fit_211_011), AICc(fit_112_011),
                 AICc(fit_111_111), AICc(fit_111_012)), 2),
  BIC  = round(c(BIC(manual_fit), BIC(fit_111_011), BIC(fit_211_011), BIC(fit_112_011),
                 BIC(fit_111_111), BIC(fit_111_012)), 2)
)

extended_candidates$ΔAIC <- round(extended_candidates$AIC - min(extended_candidates$AIC), 2)
extended_candidates[order(extended_candidates$AIC), ] %>%
  kable(row.names = FALSE, caption = "AIC Comparison — All SARIMA Candidates") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

**SARIMA(1,1,1)(0,1,1)[12]** achieves the lowest AIC across all candidates (ΔAIC = 0). The additional seasonal parameters in SARIMA(1,1,1)(1,1,1)[12] and SARIMA(1,1,1)(0,1,2)[12] are not statistically significant and do not improve either the AIC or the Ljung-Box test, confirming that the extra complexity is not justified. This model is therefore selected as the final SARIMA model. 

---

### 3.5 ETS Model

As an alternative to SARIMA, we fit an ETS (Error–Trend–Seasonality) model using `ets()`, which automatically selects the best combination of error, trend, and seasonal components by minimising the AICc.

```{r}
ets_fit <- ets(ts_train_t)
summary(ets_fit)
```

The model selected is **ETS(`r ets_fit$method`)**. Because the series was log-transformed before fitting, the multiplicative variance has already been stabilised, so ETS operates in the additive error framework on the transformed scale.

#### Residual diagnostics

```{r}
checkresiduals(ets_fit)
```

The residuals fluctuate around zero with no visible trend or seasonal structure and the histogram is approximately bell-shaped. The Ljung-Box test result should be interpreted in context: ETS(A,Ad,A) uses a damped additive trend, which — as the forecast evaluation in Section 4 reveals — ultimately fails to follow the persistent downward trend present in this series.

#### AIC comparison — SARIMA vs ETS

Note: AIC values from `Arima()` and `ets()` are computed on the same log-transformed training set and are therefore directly comparable for model selection purposes.

```{r}
all_candidates <- data.frame(
  Model = c(
    "SARIMA(0,1,1)(0,1,1)[12]",
    "SARIMA(1,1,1)(0,1,1)[12]",
    "SARIMA(2,1,1)(0,1,1)[12]",
    "SARIMA(1,1,2)(0,1,1)[12]",
    "SARIMA(1,1,1)(1,1,1)[12]",
    "SARIMA(1,1,1)(0,1,2)[12]",
    ets_fit$method
  ),
  AIC  = round(c(AIC(manual_fit), AIC(fit_111_011), AIC(fit_211_011), AIC(fit_112_011),
                 AIC(fit_111_111), AIC(fit_111_012), AIC(ets_fit)), 2),
  AICc = round(c(AICc(manual_fit), AICc(fit_111_011), AICc(fit_211_011), AICc(fit_112_011),
                 AICc(fit_111_111), AICc(fit_111_012), AICc(ets_fit)), 2),
  BIC  = round(c(BIC(manual_fit), BIC(fit_111_011), BIC(fit_211_011), BIC(fit_112_011),
                 BIC(fit_111_111), BIC(fit_111_012), BIC(ets_fit)), 2)
)

all_candidates$ΔAIC <- round(all_candidates$AIC - min(all_candidates$AIC), 2)
all_candidates[order(all_candidates$AIC), ] %>%
  kable(row.names = FALSE, caption = "AIC Comparison — All Candidates (SARIMA and ETS)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Among the SARIMA candidates, **SARIMA(1,1,1)(0,1,1)[12]** achieves the lowest AIC. ETS(A,Ad,A) has a substantially higher AIC on the log-transformed scale, already suggesting weaker in-sample fit. Both are carried forward to the forecast evaluation in Section 4, where out-of-sample accuracy on the test set determines the final selection.

---

## 4. Forecast Evaluation

Both finalist models are used to produce forecasts over the held-out test period. All predictions are back-transformed to the original (calendar-adjusted, thousand tonnes) scale before computing accuracy metrics.

### 4.1 Generating Test-Period Forecasts

```{r}
h <- length(ts_test)

sarima_fc <- forecast(fit_111_011, h = h)
ets_fc    <- forecast(ets_fit,     h = h)

# Back-transform helper (mirrors the transformation applied in Section 2.8)
inv_transform <- function(x) {
  if (!transformed) return(x)
  if (abs(lambda) < 0.25) return(exp(x))
  InvBoxCox(x, lambda)
}

sarima_mean <- inv_transform(as.numeric(sarima_fc$mean))
sarima_lo80 <- inv_transform(as.numeric(sarima_fc$lower[, 1]))
sarima_hi80 <- inv_transform(as.numeric(sarima_fc$upper[, 1]))
sarima_lo95 <- inv_transform(as.numeric(sarima_fc$lower[, 2]))
sarima_hi95 <- inv_transform(as.numeric(sarima_fc$upper[, 2]))

ets_mean <- inv_transform(as.numeric(ets_fc$mean))
ets_lo80 <- inv_transform(as.numeric(ets_fc$lower[, 1]))
ets_hi80 <- inv_transform(as.numeric(ets_fc$upper[, 1]))
ets_lo95 <- inv_transform(as.numeric(ets_fc$lower[, 2]))
ets_hi95 <- inv_transform(as.numeric(ets_fc$upper[, 2]))
```

### 4.2 Visual Comparison

```{r fig.height=5}
test_times <- as.numeric(time(ts_test))

# Last 3 years of training for visual context
tail_start  <- time(ts_adj)[max(1, train_size - 35)]
train_tail  <- window(ts_adj, start = tail_start, end = time(ts_adj)[train_size])
df_train    <- data.frame(time = as.numeric(time(train_tail)), value = as.numeric(train_tail))

df_test_obs <- data.frame(time = test_times, value = as.numeric(ts_test))

df_sarima <- data.frame(
  time  = test_times,
  mean  = sarima_mean, lo80 = sarima_lo80, hi80 = sarima_hi80,
  lo95  = sarima_lo95, hi95 = sarima_hi95,
  model = "SARIMA(1,1,1)(0,1,1)[12]"
)

df_ets <- data.frame(
  time  = test_times,
  mean  = ets_mean, lo80 = ets_lo80, hi80 = ets_hi80,
  lo95  = ets_lo95, hi95 = ets_hi95,
  model = ets_fit$method
)

df_fc <- rbind(df_sarima, df_ets)

ggplot() +
  geom_line(data = df_train,    aes(x = time, y = value), colour = "grey55", linewidth = 0.6) +
  geom_ribbon(data = df_fc, aes(x = time, ymin = lo95, ymax = hi95, fill = model), alpha = 0.15) +
  geom_ribbon(data = df_fc, aes(x = time, ymin = lo80, ymax = hi80, fill = model), alpha = 0.20) +
  geom_line(data = df_fc,   aes(x = time, y = mean, colour = model), linewidth = 0.85) +
  geom_line(data = df_test_obs, aes(x = time, y = value), colour = "black", linewidth = 0.85) +
  labs(
    title  = "Test-Period Forecasts vs Observed Values",
    subtitle = "Black line = observed; shaded bands = 80% and 95% prediction intervals",
    x = "Year", y = "Thousand tonnes",
    colour = "Model", fill = "Model"
  ) +
  theme_minimal()
```

### 4.3 Accuracy Metrics

Point forecast accuracy is evaluated on the original (calendar-adjusted) scale using RMSE, MAE, and MAPE.

```{r}
sarima_ts_orig <- ts(sarima_mean, start = start(ts_test), frequency = 12)
ets_ts_orig    <- ts(ets_mean,    start = start(ts_test), frequency = 12)

acc_sarima <- accuracy(sarima_ts_orig, ts_test)
acc_ets    <- accuracy(ets_ts_orig,    ts_test)

acc_table <- data.frame(
  Model = c("SARIMA(1,1,1)(0,1,1)[12]", ets_fit$method),
  RMSE  = round(c(acc_sarima[1, "RMSE"], acc_ets[1, "RMSE"]), 3),
  MAE   = round(c(acc_sarima[1, "MAE"],  acc_ets[1, "MAE"]),  3),
  MAPE  = round(c(acc_sarima[1, "MAPE"], acc_ets[1, "MAPE"]), 2)
)

acc_table[order(acc_table$RMSE), ] %>%
  kable(row.names = FALSE, caption = "Forecast Accuracy on the Test Set (original scale, thousand tonnes)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
best_idx   <- which.min(acc_table$RMSE)
best_model <- acc_table$Model[best_idx]
cat("Best model by RMSE:", best_model, "\n")
```

**SARIMA(1,1,1)(0,1,1)[12]** outperforms ETS(A,Ad,A) on every metric: RMSE 5.18 vs 9.90, MAE 4.16 vs 8.78, MAPE 8.05% vs 16.83% — roughly half the error in all cases. The gap is explained by a fundamental failure of the ETS model: ETS(A,Ad,A) employs a damped additive trend that, once the damping parameter pulls the trend toward zero, projects production at around 65–70 thousand tonnes — the opposite of what the data shows. The series underwent a persistent, monotonic decline from ~2008 to 2026, and the ETS framework could not extrapolate this structural break over the 94-month test horizon. SARIMA also overestimates the level by approximately 5–10 thousand tonnes on average (the blue forecast line sits above the black observed line throughout), because the decline accelerated after 2019 in a way not fully captured by the training period. Nevertheless, SARIMA correctly tracks the seasonal pattern and direction of change, making it the clear winner. It is selected as the final model.

---

### 4.4 Final Forecast

The winning model is re-fitted to the **full** calendar-adjusted series (`ts_adj`) and used to produce a 24-month-ahead forecast (March 2026 – February 2028).

```{r}
# Transform the full series with the same lambda used in training
ts_adj_t <- if (abs(lambda) < 0.25) log(ts_adj) else if (!transformed) ts_adj else BoxCox(ts_adj, lambda)

# Refit SARIMA(1,1,1)(0,1,1)[12] on the full dataset
final_fit <- Arima(ts_adj_t, order = c(1, 1, 1), seasonal = list(order = c(0, 1, 1), period = 12))
summary(final_fit)
```

```{r fig.height=5}
final_fc <- forecast(final_fit, h = 24)

final_mean <- inv_transform(as.numeric(final_fc$mean))
final_lo80 <- inv_transform(as.numeric(final_fc$lower[, 1]))
final_hi80 <- inv_transform(as.numeric(final_fc$upper[, 1]))
final_lo95 <- inv_transform(as.numeric(final_fc$lower[, 2]))
final_hi95 <- inv_transform(as.numeric(final_fc$upper[, 2]))

last_time    <- as.numeric(time(ts_adj)[length(ts_adj)])
future_times <- seq(last_time + 1/12, by = 1/12, length.out = 24)

df_hist   <- data.frame(time = as.numeric(time(ts_adj)), value = as.numeric(ts_adj))
df_future <- data.frame(
  time  = future_times,
  mean  = final_mean,
  lo80  = final_lo80, hi80 = final_hi80,
  lo95  = final_lo95, hi95 = final_hi95
)

ggplot() +
  geom_line(data = df_hist,   aes(x = time, y = value), colour = "grey45", linewidth = 0.55) +
  geom_ribbon(data = df_future, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.18) +
  geom_ribbon(data = df_future, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.30) +
  geom_line(data = df_future,   aes(x = time, y = mean),  colour = "steelblue", linewidth = 1) +
  labs(
    title    = "24-Month Ahead Forecast — SARIMA(1,1,1)(0,1,1)[12]",
    subtitle = "Shaded bands: 80% (darker) and 95% (lighter) prediction intervals",
    x = "Year", y = "Thousand tonnes"
  ) +
  theme_minimal()
```

```{r}
# Tabulate the forecast values
forecast_months <- seq.Date(as.Date("2026-03-01"), by = "month", length.out = 24)

forecast_table <- data.frame(
  Month    = format(forecast_months, "%Y-%m"),
  Forecast = round(final_mean, 2),
  Lo_80    = round(final_lo80, 2),
  Hi_80    = round(final_hi80, 2),
  Lo_95    = round(final_lo95, 2),
  Hi_95    = round(final_hi95, 2)
)

forecast_table %>%
  kable(col.names = c("Month", "Forecast", "Lower 80%", "Upper 80%", "Lower 95%", "Upper 95%"),
        caption = "24-Month Ahead Point Forecasts and Prediction Intervals (thousand tonnes)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The 24-month forecast continues the downward trend captured in the full training series, projecting production declining from ~58 thousand tonnes in March 2026 to ~45 thousand tonnes in the summer months, with the expected winter recovery (December 2026: ~52 thousand tonnes, January 2027: ~54). By February 2028 the model projects ~55 thousand tonnes, within the range observed in 2024–2026. The widening prediction intervals reflect increasing uncertainty further into the future, as is typical of SARIMA models, but the 95% bands remain physically plausible given the historical range of the series.
