---
title: "Series temporais Projeto"
output: html_notebook
---
Necessary packages
```{r}
#install.packages("readxl")
#install.packages("lubridate")

library(readxl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forecast)
```

## Data loading and preparation

```{r}
dados <- read.csv("leite.csv")
dados
```

Seleção das colunas relevantes

```{r}
dadosSelected <- select(dados, TIME_PERIOD, OBS_VALUE)
dadosSelected <- dadosSelected[-c(1,2, 3), ] #eliminação de meses vazios no inicio
head(dadosSelected)
```

```{r}
ts_data <- ts(dadosSelected$OBS_VALUE, start = c(1987, 1), frequency = 12)

```

## Exploratory Analysis

```{r}
dadosSelected$OBS_VALUE <- as.numeric(gsub(",", ".", dadosSelected$OBS_VALUE))
boxplot(dadosSelected$OBS_VALUE, main = "Boxplot of milk production in Portugal", col = "lightblue")

```

No presence of outliers

```{r}
autoplot(ts_data,
         ylab = "Thousand tonnes",
         xlab = "Year",
         main = "Monthly drinking milk production in Portugal (1987-2026)")
```

Transformation to the month lenght 
```{r}
dados_ajustados <- dadosSelected %>%
  # 1. Garante que o TIME_PERIOD seja lido como data (adicionando o dia 01)
  mutate(Data_Certa = ym(TIME_PERIOD)) %>%
  # 2. Calcula o número de dias que aquele mês específico teve
  mutate(Dias_no_Mes = days_in_month(Data_Certa)) %>%
  # 3. Faz o ajuste: valor original dividido pelo número de dias
  mutate(Valor_Por_Dia = (OBS_VALUE* (365.25/12))/ Dias_no_Mes)
```
```{r}
# Criamos a série usando a nova coluna 'Valor_Por_Dia'
ts_data_ajustada <- ts(dados_ajustados$Valor_Por_Dia, start = c(1987, 1), frequency = 12)
```
```{r}
# Desenha o gráfico comparativo
autoplot(ts_data, series = "Original (Bruto)") +
  autolayer(ts_data_ajustada , series = "Ajustado (Padronizado para 30.4 dias)") +
  labs(
    x = "Tempo", 
    y = "Valores", 
    title = "Efeito do Ajuste pelo Tamanho do Mês"
  ) +
  theme_minimal()
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
